---
title: "Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The following vignette provides an overview of the two databases stored within
`nomad`. These include a database of mobility models and an associated database
describing the different mobility datasets that were used to fit the mobility
models. Additionally, the vignette provides information on how to add new models
to the `nomad` package.

## Data Stored in `nomad`

### 1. `model_db`

The main purpose of `nomad` is store fitted mobility models for the wider research
community to use. `model_db` is a database of these fitted mobility models. 

The database is a named list, with each element representing a different fitted
model. Each model is named with a standard convention. For example, let's look at
`zmb_cdr_2020_mod_dd_exp`. 

Each of the names in the model's snake_case name refer either to the data that
the model was fit using or the type of model that was fit. For example:

Name | Description
------------- | -------------
zmb | ISO3C county code for where data was collected
cdr | Type of mobility data. cdr = Call Data Record
2020 | Year of data collection
mod | mod signifies the following labels relate to the model
dd | Which mobility model. dd = Departure-Diffusion
exp | Sub-Type of mobility model. exp = Exponential

The naming conventions help with documenting the models and enable linking
to the mobility database to query further information about the underlying
mobility data each model was fit using. All names in the snake case model name
before `mod` refer to the mobility data.

### 2. `mobility_db`

The second database in `nomad` is a table of the different mobility datasets
that are used to fit each of the models in `model_db`. The table provides
metadata about each mobility data set, which is stored to help guide users 
to the mobility data and subsequently the mobility model that may be most
suitable for their research purposes. For example, searching for data sets that
are from similar populations, or conducted at similar spatial scales. 

The table provides the following information about the mobility datasets: 

Name | Description
---- | -----------
name | Name of mobility data
country | ISO3C country code for where data was collected
date_start | Start date of data collection
date_end | End date of data collection
n | Number of data records
type | Type of mobility data, e.g. call data records, facebook
sampling_scheme | Free text description of sample scheme
censoring | Free text description of any censoring
aggregation | Spatial scale of aggregation, e.g. admin_2
url | URL for associated publication or raw data

The `name` of the mobility data matches the start of each mobility model's data. 

## Adding new mobility models and/or data to `nomad`

When new mobility data sets are generated, users may wish to easily share the 
resultant mobility models with the wider community, without requiring users to 
re-fit the model. These models can be easily added to `nomad`. 

To ensure consistency with `nomad` and ensure data transparency and
reproducibility, please follow these steps.

#### 1. Update the mobility database 

First, we need to describe the mobility dataset that the models have been fit to. 
In `nomad` we store a database (`nomad::mobility_db`), which describes all the 
mobility datasets that have been used to fit the models in `nomad`. If you have 
generated a new mobility dataset, a description of this dataset should be added
to `nomad::mobility_db`. This can be achieved as follows:

1. Fork the `nomad` repository and make a new branch
2. Go into `data-raw/` and add a new row in `mobility_db.csv` to describe your new mobility data set
3. Run the `mobility_db_load.R` script to generate the `mobility_db` R object
4. Save the new database object with `usethis::use_data(mobility_db, overwrite = TRUE)`.

#### 2. Add your mobility model to `nomad`

To ensure transparency and reproducibility, it is good practice to provide the
code used to generate the data stored in R packages (where possible). For all 
mobility models stored in `nomad`, we need to provide the code used to generate
these. 

The structure chosen in `nomad` is to include this code in `data-raw/models`. In
this location there is a directory for each mobility dataset. Each directory 
includes a script that demonstrates how the mobility data was processed and used
to fit mobility models before adding them to `model_db`.

To include your new mobility models, please follow these steps:

1. Create a new directory in `data-raw/models`, which is named after the mobility
dataset that you have added to `mobility_db`.
2. Create an R script in this directory and name it after the mobility dataset. 
3. Edit the R script to show how the mobility dataset is loaded and used to fit 
different mobility models before adding them to `nomad::model_db`. 
4. Please include all raw data used in the R script in the same directory.

An example as used for the Zambia 2020 Call Data Records (`data-raw/models/zmb_cdr_2020`)
is shown below:

```{r, eval = FALSE}
## Code to add zmb_cdr_2020 models to database

# Read in data sets
D <- readRDS(file.path(here::here(), "data-raw/models/zmb_cdr_2020/dist_data.rds"))
M <- readRDS(file.path(here::here(), "data-raw/models/zmb_cdr_2020/OD_data_diag.rds"))
N <- readRDS(file.path(here::here(), "data-raw/models/zmb_cdr_2020/pop_data.rds"))

# Convert to matrices
D <- as.matrix(D)
M <- as.matrix(M)

# Set seed for reproducibility
set.seed(123L)

# Create mobility model
mod <- mobility::mobility(
  data = list("D" = D, "M" = M, "N" = N),
  model = "departure-diffusion",
  type = "exp",
  DIC = TRUE)

# Convert to nomad model
zmb_cdr_2020_mod_dd_exp <- nomad::nomad_model(
  model = mod,
  data_name = "zmb_cdr_2020"
)

# Add to the model database
model_db <- nomad:::add_model_to_db(zmb_cdr_2020_mod_dd_exp)

# Update the package model database
usethis::use_data(model_db, overwrite = TRUE)
```

If the model you have built uses data that you cannot share publicly, please
still include the code you did use and the datasets that you can share (e.g. 
population and distance matrices data) and after you have created the model
you can remove the data using `nomad_remove_model_data()`. 

If the data is removed, however, the methods for checking and returning the
residuals of the `mobility::mobility()` model will not work. `nomad` provides
a solution to this that allows for the model performance statistics and plot
to be saved when the underlying data is removed with the argument
`save_model_checks`. For example, in the `nomad` package, we include models
that are based on Facebook data that can not be shared publicly. However, we 
can save the model check summaries in `nomad` as follows (this example can 
be seen in `data-raw/models/zmb_fb_2020`).

```{r, eval = FALSE}
# First define where we are saving the model_check.png to in data-raw
model_check_path <- file.path(here::here(), "data-raw/models/zmb_fb_2020/model_check.png")

# Now remove the data but save_model_checks
zmb_fb_2020_mod_grav_exp <- nomad_remove_model_data(
  zmb_fb_2020_mod_grav_exp,
  M = TRUE,
  save_model_checks = TRUE,
  # the following arguments are passed to png to save the model check plot
  filename = model_check_path,
  width = 8, 
  height = 4, 
  units = "in", 
  res = 300
  )

```

This will now save the model check statistics in the model object, which can be 
accessed with `zmb_fb_2020_mod_grav_exp$get_check_res()` and save the check plot
to file, specified with `filename`. To then make this plot object available to 
other users to see, we then need to copy it into `inst/extdata`. Please follow
the example laid out below (also in `data-raw/models/zmb_fb_2020`):

```{r, eval = FALSE}
# Now copy the model check plot to the package external directory
copy_loc <- file.path(
  here::here(),
  "inst/extdata/model_checks",
  file.path(basename(dirname(model_check_path)), basename(model_check_path))
)
dir.create(dirname(copy_loc), recursive = TRUE, showWarnings = FALSE)
file.copy(model_check_path, copy_loc)
```

This means the check plots are available for other users to view. In `nomad`, 
when users wish to `check()` model objects, it will return the same outputs now
as if the data was shared. However, the `residuals()` function will not return
an output (as the end user could combine the outputs of `predict()` and 
`residuals()` to recover the underlying data)

#### 3. Incorporate changes and submit pull request

1. Check that the changes have not impacted the package `devtools::check()` (Ctrl+Shift+E)
2. Commit your changes and push changes to your fork on Github
3. Make a pull request with your new model and tag one of the maintainers of 
`nomad` (@OJWatson).


