---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>"
)
source(system.file("rmarkdown", "folding.R", package = "nomad"))
```

`nomad` stores fitted mobility models for infectious disease modelling. The
usual workflow is:

1. identify a plausible source mobility dataset;
2. inspect the fitted models available for that dataset;
3. prepare a distance matrix and population vector for the prediction setting;
4. predict a mobility matrix;
5. check warnings and sensitivity to alternative models.

```{r setup}
library(nomad)
```

## Available Data And Models

The mobility catalogue describes the source datasets used to fit models in the
package. It is the first thing to check because a model fitted to the wrong
country, period, or spatial scale can be more misleading than helpful.

```{r}
mobility_table()
```

Check `has_fitted_models` before selecting a dataset: some catalogue entries
are awaiting data and have no model yet. The model catalogue below lists the
actual fitted variants, along with their source metadata and fit statistics.

```{r}
model_table(country = "ZMB")
```

## Inspect A Model

The model profile summarises the scale and period of the source data. This is
the quickest way to check whether a model is plausible for a prediction task.

```{r}
model <- model_db$zmb_fb_2025_mod_grav_exp_norm
model_profile(model)
```

Model-check statistics and plots are available even when the underlying private
mobility matrix cannot be bundled.

```{r}
check(model, plots = FALSE)
plot(model)
```

## Prepare A Small Prediction Setting

Predictions use the same data structure as `mobility::predict()`: a distance
matrix `D` and either one population vector `N` or origin/destination vectors
`N_orig` and `N_dest`.

Here we make a small, named toy setting. The values are not intended to be real
Zambia data; they are just enough to show the shape of a prediction request.

```{r, fold=TRUE, fold_label="Show how to construct the example regions"}
places <- data.frame(
  name = c("Lusaka", "Kabwe", "Ndola", "Kitwe", "Livingstone"),
  x = c(0, 1.2, 2.8, 2.6, -1.4),
  y = c(0, 1.0, 2.7, 2.4, -2.7),
  population = c(2500000, 230000, 475000, 690000, 170000)
)

D <- as.matrix(dist(places[, c("x", "y")])) * 120
dimnames(D) <- list(places$name, places$name)
N <- setNames(places$population, places$name)
newdata <- list(D = D, N = N)
```

## Make A Prediction

This toy request is smaller than the model source data, so `nomad` warns. That
is deliberate: warnings are part of the workflow, not just noise to suppress.

```{r}
M <- predict(model, newdata = newdata, unit = "km")
round(M / 1000, 1)
```

Rows are origins and columns are destinations. These displayed values are in
thousands of predicted trips over 21 days, not probabilities. A person can
contribute more than one trip. The names on the matrix and population vector
must identify the same regions; nomad aligns named populations before prediction.

For a real analysis, the next step would be to decide whether the warning is
acceptable, whether a better-matched model exists, or whether the prediction
regions should be changed.

## Use Source Data When You Want A Clean Demonstration

The fitted model also contains public covariates from the source setting. Reusing
those covariates is useful for checking package mechanics without introducing a
scale mismatch.

```{r}
source_data <- model$get_model()$data[c("D", "N_orig", "N_dest")]
M_source <- predict(model, newdata = source_data)
M_source[1:5, 1:5]
```

If a prediction should represent a different time period, supply `duration` in
days or a `date_start`/`date_end` pair. The new Facebook models represent 21-day
trip counts; collection dates are a separate field. Scaling assumes a constant
trip rate. Original Zambia models with an unknown prediction period can only
return unscaled predictions, with a warning.

```{r}
cdr_model <- model_db$zmb_fb_2025_mod_dd_exp
cdr_data <- cdr_model$get_model()$data[c("D", "N_orig", "N_dest")]

M_30 <- predict(cdr_model, newdata = cdr_data, duration = 30)
attr(M_30, "duration_scale")
```

## Compare Alternatives

When more than one model is plausible, use model-fit statistics to make the
comparison explicit. `model_weights()` gives simple candidate weights from a
chosen metric; users should still decide whether the source data are relevant.

```{r}
weights <- model_weights(data_name = "zmb_fb_2025", metric = "RMSE")
weights[, c("name", "RMSE", "weight")]
```

`nomad_ensemble()` combines predictions from several models using explicit
weights. This example compares two families fitted to the same 21-day dataset.
Each is predicted for the same target regions, so the outputs can be combined.
The weights sum to one; 0.5 gives each prediction equal influence.

```{r}
ensemble <- nomad_ensemble(
  list(cdr = cdr_model, facebook = model),
  weights = c(cdr = 0.5, facebook = 0.5)
)
M_ensemble <- predict(ensemble, newdata = newdata, unit = "km")
round(sum(M_ensemble))
```

## Simulated Predictions

With `nsim > 1`, predictions are stochastic draws. `nomad` can summarise and
select representative lower, median, and upper matrices.

```{r}
draws <- predict(model, newdata = source_data, nsim = 10, seed = 1)
uncertainty_summary(draws, metrics = c("total", "between_region"))
```

The "Selecting a mobility model" vignette gives more detail on warnings,
uncertainty summaries, prediction differences, and ensembles. The
"Identifying a mobility model for an outbreak" vignette shows how model choice
can be linked to an epidemiological outcome rather than only to mobility fit
metrics.
