---
title: "Models and Mobility Data"
output: rmarkdown::html_document
resource_files:
  - model_checks/
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>", fold = TRUE)
source(system.file("rmarkdown", "folding.R", package = "nomad"))
library(nomad)
```

<div id="catalogue-page"></div>

This page describes the mobility datasets and fitted models available in `nomad`.
Search the tables for a country or dataset, explore its coverage on the map, and
inspect the model fit plots before choosing a model for your analysis. Expand the
code panels to use the same tables and plots in R.

## Mobility Data

Start by finding data that are representative of your setting: the country,
spatial aggregation and collection period are useful starting points. Search
across the table, click a column heading to sort it, or expand a row for sampling
and censoring details. Datasets without fitted models remain visible and are
marked in the availability column.

```{r datasets, fold_label="Show how to browse mobility datasets in R"}
datasets <- mobility_table()
reactable::reactable(
  datasets[, c("name", "country", "type", "n", "aggregation", "date_start",
                "date_end", "has_fitted_models")],
  searchable = TRUE, resizable = TRUE, wrap = FALSE,
  bordered = TRUE, highlight = TRUE,
  showPageSizeOptions = TRUE, pageSizeOptions = c(4, 8, 12, 24), defaultPageSize = 8,
  defaultSorted = list(date_start = "desc"),
  defaultColDef = reactable::colDef(minWidth = 100),
  theme = reactable::reactableTheme(highlightColor = "#eaf4f8"),
  columns = list(
    name = reactable::colDef(name = "Dataset", sticky = "left", minWidth = 175,
                             style = list(borderRight = "1px solid #ccd5da")),
    country = reactable::colDef(name = "ISO3", minWidth = 65),
    type = reactable::colDef(name = "Data", minWidth = 130),
    n = reactable::colDef(name = "Records", minWidth = 140, format = reactable::colFormat(digits = 0, separators = TRUE)),
    aggregation = reactable::colDef(name = "Spatial scale"),
    date_start = reactable::colDef(name = "Start"),
    date_end = reactable::colDef(name = "End"),
    has_fitted_models = reactable::colDef(name = "Fitted models", minWidth = 140,
                                          cell = function(x) if (x) "Yes" else "Not yet")
  ),
  details = function(index) {
    row <- datasets[index, ]
    period <- if (is.na(row$prediction_period_days)) "Unknown" else
      paste(row$prediction_period_days, "days")
    htmltools::tags$div(style = "padding: 12px; white-space: normal;",
      htmltools::tags$p(htmltools::tags$b("Sampling: "), row$sampling_scheme),
      htmltools::tags$p(htmltools::tags$b("Censoring: "),
                        if (is.na(row$censoring)) "Not specified" else row$censoring),
      htmltools::tags$p(htmltools::tags$b("Prediction period: "), period),
      if (!is.na(row$publication)) htmltools::tags$a("Publication", href = row$publication)
    )
  }
)
```

The start and end dates describe data collection. The period represented by a
prediction can differ; expand a dataset's row to see that period separately.

## Where are the data from?

Switch between Facebook and call data records to see the countries covered by
each source. Hover over a country to highlight it, then click for dataset details.
Colour shows the number of records in the latest catalogued dataset for each
country and source; definitions and collection periods vary between datasets.
The source-region selector adds administrative centroids for the newer Facebook
datasets.

```{r coverage-map, fold_label="Show how to make the coverage map in R"}
world <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
locations <- readRDS(system.file("extdata", "source_locations.rds", package = "nomad"))
map_data <- datasets[order(datasets$date_end, decreasing = TRUE, na.last = TRUE), ]
map_data <- map_data[!duplicated(paste(map_data$country, map_data$type)), ]
world$country <- world$iso_a3
world$country[world$country == "-99"] <- world$adm0_a3[world$country == "-99"]
coverage <- merge(world[, c("country", "name_long")], map_data, by = "country")
coverage$log_records <- log10(coverage$n)
coverage$popup <- paste0(
  "<b>", coverage$name_long, "</b><br>", coverage$name,
  "<br>", coverage$date_start, " to ", coverage$date_end,
  "<br>Records: ", format(round(coverage$n), big.mark = ",", scientific = FALSE, trim = TRUE),
  "<br>Spatial scale: ", coverage$aggregation,
  "<br>Fitted models: ", ifelse(coverage$has_fitted_models, "available", "not yet available")
)
palette <- leaflet::colorBin("YlOrBr", domain = coverage$log_records, bins = 5)
map <- leaflet::leaflet(coverage, width = "100%", height = 550)
map <- leaflet::addTiles(map)
for (source in c("facebook", "call data record")) {
  layer <- coverage[coverage$type == source, ]
  group <- if (source == "facebook") "Facebook" else "Call data records"
  map <- leaflet::addPolygons(map, data = layer, group = group,
    fillColor = ~palette(log_records), fillOpacity = 0.75,
    color = "#8c949b", weight = 1, popup = ~popup, label = ~name_long,
    highlightOptions = leaflet::highlightOptions(weight = 3, color = "#2c3e50",
                                                  bringToFront = TRUE))
}
map <- leaflet::addLayersControl(map, baseGroups = c("Facebook", "Call data records"),
                                 options = leaflet::layersControlOptions(collapsed = FALSE))
map <- leaflet::hideGroup(map, "Call data records")
map <- leaflet::addLegend(map, pal = palette, values = ~log_records,
                          title = "Records (log10)", position = "bottomleft")
ids <- sort(unique(locations$data_name))
for (id in ids) {
  points <- locations[locations$data_name == id, ]
  map <- leaflet::addCircleMarkers(map, data = points, lng = ~lon, lat = ~lat,
    radius = 3, color = "#2166ac", fillOpacity = 0.85, group = id,
    popup = ~htmltools::htmlEscape(label))
}
map <- leaflet::hideGroup(map, ids)
map <- leaflet::fitBounds(map, lng1 = -26, lat1 = -36, lng2 = 63, lat2 = 39)
htmlwidgets::onRender(map, '
function(el, x, ids) {
  var map = this;
  var control = L.control({position: "topright"});
  control.onAdd = function() {
    var label = L.DomUtil.create("label");
    label.style.cssText = "background: white; padding: 8px; border-radius: 4px";
    label.textContent = "Source regions: ";
    var select = document.createElement("select");
    select.setAttribute("aria-label", "Source dataset");
    var none = document.createElement("option");
    none.value = ""; none.textContent = "Country coverage";
    select.appendChild(none);
    ids.forEach(function(id) {
      var option = document.createElement("option");
      option.value = id; option.textContent = id;
      select.appendChild(option);
    });
    select.onchange = function() {
      ids.forEach(function(id) {
        var group = map.layerManager.getLayerGroup(id);
        if (group) map.removeLayer(group);
      });
      var selected = map.layerManager.getLayerGroup(select.value);
      if (selected && select.value) {
        map.addLayer(selected);
        map.fitBounds(L.featureGroup(selected.getLayers()).getBounds());
      } else {
        map.fitBounds([[-36, -26], [39, 63]]);
      }
    };
    label.appendChild(select);
    L.DomEvent.disableClickPropagation(label);
    L.DomEvent.disableScrollPropagation(label);
    return label;
  };
  control.addTo(map);
}', data = ids)
```

## Model Performance

Search for models fitted to a dataset you identified above, or filter by country,
model family or type. Sort the fit statistics by clicking their headings. Compare
fit errors within a source dataset, and use the plots below to see where models
fit well or poorly. Missing DIC means that DIC weighting is unavailable for that fit.

```{r fitted-models, fold_label="Show how to compare fitted models in R"}
models <- model_table()
reactable::reactable(
  models[, c("name", "country", "mobility_model", "model_type", "RMSE", "MAPE",
              "R2", "DIC", "max_rhat", "supports_newdata")],
  searchable = TRUE, filterable = TRUE, resizable = TRUE, wrap = FALSE,
  bordered = TRUE, highlight = TRUE,
  defaultSorted = list(R2 = "desc"),
  showPageSizeOptions = TRUE, pageSizeOptions = c(4, 8, 12, 24), defaultPageSize = 8,
  theme = reactable::reactableTheme(highlightColor = "#eaf4f8"),
  defaultColDef = reactable::colDef(minWidth = 90),
  columns = list(
    name = reactable::colDef(name = "Model", sticky = "left", minWidth = 260,
                             style = list(borderRight = "1px solid #ccd5da")),
    country = reactable::colDef(name = "ISO3", minWidth = 65),
    mobility_model = reactable::colDef(name = "Family", minWidth = 150),
    model_type = reactable::colDef(name = "Type", minWidth = 120),
    RMSE = reactable::colDef(format = reactable::colFormat(digits = 1, separators = TRUE)),
    MAPE = reactable::colDef(format = reactable::colFormat(digits = 2)),
    R2 = reactable::colDef(format = reactable::colFormat(digits = 3)),
    DIC = reactable::colDef(format = reactable::colFormat(digits = 0, separators = TRUE)),
    max_rhat = reactable::colDef(name = "Max R-hat"),
    supports_newdata = reactable::colDef(name = "New regions", minWidth = 100)
  )
)
```

A high stored R-hat indicates that the fit needs convergence review; a value near
one alone does not establish suitability for your application. For a chosen
model, `summary(model)` provides parameter-level diagnostics. Basic and finite
radiation models can only predict for their original regions and populations.

## Model Fit Checks

Summary statistics can hide systematic differences between observed and fitted
flows. Choose a model below to inspect its saved diagnostic plot. The first panel
compares observed and predicted flows; the residual plot helps identify departures
that an overall error measure can miss.

```{r gallery-assets, include=FALSE}
plot_dir <- "model_checks"
dir.create(plot_dir, showWarnings = FALSE)
for (id in names(model_db)) {
  source <- nomad:::check_plot_file_path(model_db[[id]])
  destination <- file.path(plot_dir, paste0(id, ".png"))
  if (file.exists(source)) {
    file.copy(source, destination, overwrite = TRUE)
  } else {
    grDevices::png(destination, width = 1200, height = 600, res = 150)
    plot(model_db[[id]])
    grDevices::dev.off()
  }
}
```

```{r gallery, echo=FALSE, results="asis"}
selected <- "zmb_fb_2025_mod_grav_exp_norm"
options <- lapply(names(model_db), function(id) {
  htmltools::tags$option(value = id, selected = if (id == selected) "selected" else NULL, id)
})
htmltools::tags$div(
  htmltools::tags$label("Model: ", `for` = "plotSelect"),
  htmltools::tags$select(id = "plotSelect", style = "max-width:100%; margin-bottom:1em;",
    onchange = "document.getElementById('plotImage').src = 'model_checks/' + this.value + '.png'; document.getElementById('plotImage').alt = 'Fit diagnostics for ' + this.value;",
    options),
  htmltools::tags$img(id = "plotImage", class = "catalogue-plot",
    src = paste0("model_checks/", selected, ".png"),
    alt = paste("Fit diagnostics for", selected))
)
```

```{r fit-in-r, eval=FALSE, fold_label="Show how to inspect the same fit in R"}
model <- model_db$zmb_fb_2025_mod_grav_exp_norm
check(model, plots = FALSE)
plot(model)
summary(model)
```

Continue to [Selecting a mobility model](selecting-a-model.html) for a worked
comparison of predictions, uncertainty and ensembles.
