---
title: "Models and Mobility Data"
output:
  rmarkdown::html_vignette:
    toc: false
vignette: >
  %\VignetteIndexEntry{Models and Mobility Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>"
)
source(system.file("rmarkdown", "folding.R", package = "nomad"))
```

```{r setup}
library(nomad)
```

This vignette describes the mobility data sources and fitted models currently
available in `nomad`. Use it to identify source data that are plausible for
your analysis before moving to the more detailed model-selection vignettes.

## Mobility Data

`mobility_table()` returns the metadata catalogue for source mobility datasets.
The most important fields for model choice are usually country, data type,
collection period, sample size, and spatial aggregation. `has_fitted_models`
indicates whether a dataset has any selectable model; FALSE entries are retained
as catalogue information. The original Zambia prediction periods are unknown,
while the new Facebook matrices represent average 21-day trip counts.

```{r}
mobility_table()
```

For the package website, the same information is shown as a searchable table.

```{r, echo = FALSE}
mobility_display <- mobility_table()

reactable::reactable(
  mobility_display,
  searchable = TRUE,
  bordered = TRUE,
  highlight = TRUE,
  defaultPageSize = 8,
  columns = list(
    name = reactable::colDef(name = "Dataset", sticky = "left", minWidth = 150),
    country = reactable::colDef(name = "ISO3", minWidth = 70),
    n = reactable::colDef(
      name = "Records",
      cell = function(value) {
        if (is.na(value)) {
          return(NA)
        }
        format(value, big.mark = ",")
      }
    ),
    has_fitted_models = reactable::colDef(name = "Fitted models in database"),
    prediction_period_days = reactable::colDef(name = "Prediction period (days)"),
    publication = reactable::colDef(
      name = "Publication",
      cell = function(value) {
        if (is.na(value)) {
          return(NA)
        }
        htmltools::tags$a(href = value, target = "_blank", as.character(value))
      }
    )
  )
)
```

## Fitted Models

`model_table()` joins fitted model metadata, source dataset metadata, and model
fit statistics.

```{r}
model_table()[, c(
  "name", "country", "data_type", "aggregation",
  "mobility_model", "model_type", "DIC", "RMSE", "MAPE", "R2", "max_rhat"
)]
```

```{r, echo = FALSE}
model_display <- model_table()

reactable::reactable(
  model_display,
  searchable = TRUE,
  bordered = TRUE,
  highlight = TRUE,
  defaultSorted = list(R2 = "desc"),
  defaultPageSize = 8,
  columns = list(
    name = reactable::colDef(name = "Model", sticky = "left", minWidth = 220),
    data_name = reactable::colDef(name = "Dataset", minWidth = 150),
    country = reactable::colDef(name = "ISO3", minWidth = 70),
    max_rhat = reactable::colDef(
      name = "Maximum R-hat",
      cell = function(value) {
        if (is.na(value)) return("Not available/applicable")
        if (value > 1.1) paste0(value, " — review fit") else as.character(value)
      }
    ),
    DIC = reactable::colDef(cell = function(value) formatC(value, format = "e", digits = 2)),
    RMSE = reactable::colDef(cell = function(value) formatC(value, format = "e", digits = 2)),
    MAPE = reactable::colDef(cell = function(value) formatC(value, digits = 2)),
    R2 = reactable::colDef(cell = function(value) formatC(value, digits = 2))
  )
)
```

Fit statistics are useful for comparing candidate models, especially when the
models were fitted to the same source dataset. They should still be interpreted
alongside source relevance: a lower error metric does not make a model suitable
for a country, time period, or spatial scale that differs from the fitted data.

The supplied summaries also contain MCMC diagnostics. `max_rhat` and
`min_n_eff` report the worst stored R-hat and effective sample size per fit.
`rhat_above_1_1` flags 111 of the 512 fits with summaries for review; basic/finite
radiation has no MCMC diagnostics. This flag is a coarse screen, not a guarantee
of convergence below that threshold. All supplied fits are retained unchanged.
Inspect `summary(model)` and the original fitting analysis before relying on a
flagged model; successful software checks do not establish statistical calibration.

## Model Profiles

`model_profile()` gives a compact summary of the source data stored in a fitted
model object. This is the quickest way to check spatial scale, population scale,
dates, and whether the original mobility data are bundled.

```{r}
model_profile(model_db$zmb_cdr_2020_mod_dd_exp)
```

```{r}
model_profile(model_db$zmb_fb_2020_mod_grav_exp)
```

## Model Fit Checks

`check()` returns model fit statistics. `plot()` draws the model-check plot when
it is available. For private-data models, `nomad` can retain these diagnostics
without bundling the underlying mobility matrix.

```{r}
check(model_db$zmb_fb_2020_mod_grav_exp, plots = FALSE)
```

```{r, fig.width = 8, fig.height = 4, dpi = 150}
plot(model_db$zmb_fb_2020_mod_grav_exp)
```

## Where To Go Next

Use the selecting-a-model vignette for warnings, uncertainty summaries,
prediction differences, and ensembles. Use the outbreak model-selection vignette
for an example of comparing mobility assumptions against a downstream
epidemiological outcome.

## Explore source locations

The map shows source countries and administrative-region centroids. Select a
new Facebook dataset to inspect its regions; a point represents an aggregated
source region, not an individual traveller or a prediction location. Country
popups distinguish the number of catalogued datasets from those with fitted
models. The metadata-only CDR entries remain in the table above.

```{r, fold_label="Show how the source map is assembled"}
locations <- readRDS(system.file("extdata", "source_locations.rds", package = "nomad"))
world <- rnaturalearth::ne_countries(scale = "small", returnclass = "sf")
world <- world[world$iso_a3 %in% mobility_display$country, ]
world$popup <- vapply(world$iso_a3, function(iso) {
  datasets <- mobility_display[mobility_display$country == iso, ]
  paste0(iso, ": ", nrow(datasets), " datasets; ",
         sum(datasets$has_fitted_models), " with fitted models")
}, character(1))
map <- leaflet::leaflet(world)
map <- leaflet::addProviderTiles(map, "CartoDB.Positron")
map <- leaflet::addPolygons(map, popup = ~popup, weight = 1, fillOpacity = 0.15)
ids <- sort(unique(locations$data_name))
for (id in ids) {
  points <- locations[locations$data_name == id, ]
  meta <- mobility_display[mobility_display$name == id, ]
  popup <- paste0(htmltools::htmlEscape(points$label), "<br>", id,
                  "<br>", meta$aggregation, "; ", meta$prediction_period_days,
                  "-day predictions; fitted models available")
  map <- leaflet::addCircleMarkers(map, data = points, lng = ~lon, lat = ~lat,
                                   radius = 3, group = id, popup = popup)
}
map <- leaflet::hideGroup(map, ids)
map <- leaflet::showGroup(map, "zmb_fb_2025")
map <- leaflet::setView(map, lng = 25, lat = -5, zoom = 3)
htmlwidgets::onRender(map, '
function(el, x, ids) {
  var map = this;
  var control = L.control({position: "topright"});
  control.onAdd = function() {
    var label = L.DomUtil.create("label");
    label.style.background = "white";
    label.style.padding = "8px";
    label.textContent = "Source dataset: ";
    var select = document.createElement("select");
    select.setAttribute("aria-label", "Source dataset");
    ids.forEach(function(id) {
      var option = document.createElement("option");
      option.value = id; option.textContent = id;
      option.selected = id === "zmb_fb_2025";
      select.appendChild(option);
    });
    select.onchange = function() {
      ids.forEach(function(id) {
        var group = map.layerManager.getLayerGroup(id);
        if (group) map.removeLayer(group);
      });
      var selected = map.layerManager.getLayerGroup(select.value);
      if (selected) {
        map.addLayer(selected);
        var layers = selected.getLayers();
        if (layers.length) map.fitBounds(L.featureGroup(layers).getBounds());
      }
    };
    label.appendChild(select);
    L.DomEvent.disableClickPropagation(label);
    return label;
  };
  control.addTo(map);
}', data = ids)
```

These centroids come from the supplied administrative geometries and match the
fitted region identifiers. The basemap needs an internet connection; packaged
metadata, centroids, fits and diagnostic plots remain available offline.
