---
title: "Identifying a mobility model for an outbreak"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Identifying a mobility model for an outbreak}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>"
)
source(system.file("rmarkdown", "folding.R", package = "nomad"))
```

Mobility fit statistics are useful, but they are not the only reason to choose
a model. In an outbreak analysis, the practical question is often whether a
mobility assumption helps explain the observed spatial pattern of cases.

This vignette uses a controlled fictional example. We create a set of regions,
simulate a spatially coupled stochastic SIR outbreak using one bundled `nomad`
model as the data-generating mobility assumption, and then ask which package
model best explains the simulated outbreak snapshot.

```{r setup}
library(nomad)
```

## A Fictional Place

The example has 24 hypothetical regions. The locations and populations are
invented, but the structure is deliberately similar to an admin-level analysis:
named regions, a population vector, and a pairwise distance matrix.

```{r}
regions <- data.frame(
  region = c(
    "Central City", "North Market", "River East", "Hill West",
    "Airport", "Lake Town", "Copper Belt", "Forest Gate",
    "South Farms", "Old Mission", "Border Post", "Rail Junction",
    "Valley Clinic", "New Estate", "Quarry", "Harbour Road",
    "Sugar Plain", "Cattle Camp", "Highlands", "Mines End",
    "University", "Bus Depot", "Garden Ward", "Wetlands"
  ),
  x = c(
    0, 0.8, 1.6, -1.1, 2.5, -2.0, 3.1, -2.6,
    0.4, -1.8, 3.8, 1.9, -0.4, 0.9, 2.8, -3.1,
    -0.9, 1.4, -2.4, 3.5, 0.2, 1.1, -1.3, -2.8
  ),
  y = c(
    0, 1.2, -0.4, 0.9, 0.7, 2.2, 2.1, -0.1,
    -1.9, -2.4, -1.3, 1.8, 2.9, -0.9, -2.5, 1.1,
    -3.2, -2.8, 3.2, 3.0, 0.4, -1.6, 1.9, -1.7
  ),
  population = c(
    1250000, 260000, 340000, 180000, 410000, 120000,
    520000, 95000, 105000, 80000, 70000, 300000,
    65000, 220000, 90000, 140000, 75000, 55000,
    85000, 160000, 380000, 240000, 130000, 60000
  )
)

D <- as.matrix(dist(regions[, c("x", "y")])) * 70
dimnames(D) <- list(regions$region, regions$region)
N <- setNames(regions$population, regions$region)
newdata <- list(D = D, N = N)

regions[1:8, ]
```

The bundled models were fitted to Zambia mobility data. Here they are used only
as alternative mobility assumptions in a fictional software example. The point
is the workflow: predict candidate mobility matrices, run the same transmission
model under each matrix, and score the resulting spatial outbreak pattern.

## Mobility Candidates From `nomad`

The candidate mobility matrices come from actual package models. We use
normalised exponential gravity and exponential departure-diffusion from the same
21-day dataset; their stored diagnostics are more suitable for a worked example
than alternatives with pronounced mixing problems. The Facebook
gravity model is the known data-generating model for this controlled example.
The Facebook departure-diffusion model is the alternative, and the ensemble weights
create intermediate candidates.

```{r}
gravity_model <- model_db$zmb_fb_2025_mod_grav_exp_norm
diffusion_model <- model_db$zmb_fb_2025_mod_dd_exp

gravity_flows <- predict(gravity_model, newdata = newdata, unit = "km")
diffusion_flows <- predict(diffusion_model, newdata = newdata, unit = "km")

candidate_prediction <- function(weight_gravity) {
  if (weight_gravity == 0) {
    return(predict(diffusion_model, newdata = newdata, unit = "km"))
  }
  if (weight_gravity == 1) {
    return(predict(gravity_model, newdata = newdata, unit = "km"))
  }

  ensemble <- nomad_ensemble(
    list(diffusion = diffusion_model, gravity = gravity_model),
    weights = c(diffusion = 1 - weight_gravity, gravity = weight_gravity)
  )
  predict(ensemble, newdata = newdata, unit = "km")
}
```

## A Spatially Coupled SIR Model

The SIR model below is intentionally compact, but it is a real stochastic
compartmental model. Each day, susceptible individuals can become infected via
local prevalence and prevalence imported through the mobility matrix. Infectious
individuals recover with a daily probability derived from the mean infectious
period.

```{r}
flow_to_coupling <- function(M) {
  M <- pmax(as.matrix(M), 0)
  diag(M) <- 0

  row_total <- rowSums(M)
  P <- M
  P[row_total > 0, ] <- P[row_total > 0, , drop = FALSE] / row_total[row_total > 0]
  P[row_total == 0, ] <- 1 / (ncol(M) - 1)
  diag(P) <- 0
  P
}

simulate_sir <- function(P,
                         N,
                         days = 45,
                         beta = 0.35,
                         gamma = 1 / 7,
                         epsilon = 0.28,
                         seed_region = "Central City",
                         seed_infections = 60) {
  S <- round(N)
  I <- setNames(rep(0, length(N)), names(N))
  R <- I
  C <- I

  I[seed_region] <- seed_infections
  S[seed_region] <- S[seed_region] - seed_infections
  C[seed_region] <- seed_infections

  p_recover <- 1 - exp(-gamma)

  for (day in seq_len(days)) {
    prevalence <- I / N
    imported <- as.numeric(P %*% prevalence)
    names(imported) <- names(N)

    lambda <- beta * ((1 - epsilon) * prevalence + epsilon * imported)
    p_infection <- pmin(1 - exp(-lambda), 1)

    new_infections <- stats::rbinom(length(N), pmax(round(S), 0), p_infection)
    recoveries <- stats::rbinom(length(N), pmax(round(I), 0), p_recover)

    S <- S - new_infections
    I <- I + new_infections - recoveries
    R <- R + recoveries
    C <- C + new_infections
  }

  C
}

mean_sir <- function(P, N, nsim = 80, seed = 1, ...) {
  set.seed(seed)
  out <- matrix(0, nrow = length(N), ncol = nsim,
                dimnames = list(names(N), NULL))
  for (i in seq_len(nsim)) {
    out[, i] <- simulate_sir(P, N, ...)
  }
  rowMeans(out)
}
```

The force of infection in region `i` is:

```{r, echo = FALSE, results = "asis"}
cat("`lambda_i = beta * ((1 - epsilon) * I_i / N_i + epsilon * imported_i)`")
```

where `imported_i` is prevalence pressure from other regions after applying the
row-normalised mobility matrix. Each origin experiences the average prevalence
of its destinations, weighted by its outgoing travel shares. This is a simplified
mixing assumption, not a simulation of travellers moving between compartments.
The row normalization discards absolute trip volumes; the daily step comes from
the SIR model, not from treating the 21-day mobility matrix as daily trips.

`S`, `I` and `R` track susceptible, infectious and recovered people. New infections
move people from S to I, and recoveries move them from I to R, preserving each
region's population. `C` accumulates infections, including the initial seed.
`beta`, `gamma` and `epsilon` remain fixed across every candidate: only mobility
changes. The `mean_sir()` helper repeats the simulation to estimate its expected
cumulative infections.

## Simulated Outbreak Data

We generate one fully observed synthetic outbreak using the gravity model.
There is no reporting fraction or additional observation model: observed cases
are cumulative infections from this stochastic realization. Candidate models will
be compared with these counts using independent simulation averages.

```{r}
truth_coupling <- flow_to_coupling(gravity_flows)
set.seed(99)
observed_cases <- simulate_sir(truth_coupling, N)
outbreak <- data.frame(
  region = names(observed_cases),
  population = as.numeric(N),
  observed_cases = observed_cases
)
outbreak[order(outbreak$observed_cases, decreasing = TRUE), ][1:12, ]
```

```{r, fig.width = 7.5, fig.height = 5}
case_plot_data <- data.frame(
  region = names(observed_cases),
  observed_cases = as.numeric(observed_cases)
)
case_plot_data <- case_plot_data[order(case_plot_data$observed_cases), ]
case_plot_data$region <- factor(case_plot_data$region, levels = case_plot_data$region)

ggplot2::ggplot(case_plot_data, ggplot2::aes(observed_cases, region)) +
  ggplot2::geom_col(fill = "#33658a", width = 0.75) +
  ggplot2::scale_x_continuous(
    labels = function(x) format(round(x), big.mark = ",")
  ) +
  ggplot2::labs(
    title = "Synthetic outbreak snapshot",
    subtitle = "Observed cumulative cases by fictional region",
    x = "Observed cumulative cases",
    y = NULL
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    panel.grid.major.y = ggplot2::element_blank(),
    plot.title.position = "plot"
  )
```

## Score Candidate Mobility Models

For each candidate, we predict a mobility matrix with `nomad`, convert it to the
SIR coupling matrix, run repeated stochastic SIR simulations, and compare the
mean final cumulative infections with the observed outbreak.

We compare cumulative infection counts directly, without adjusting each
candidate's total to match the observations. Log RMSE gives large and small
outbreaks a more balanced contribution than raw-count RMSE. A smaller value means
closer agreement for this example; it is not a likelihood or a universal criterion.

```{r}
log_rmse <- function(observed, predicted) {
  sqrt(mean((log1p(observed) - log1p(predicted))^2))
}

score_candidate <- function(weight_gravity) {
  flows <- candidate_prediction(weight_gravity)
  P <- flow_to_coupling(flows)
  predicted_infections <- mean_sir(
    P,
    N,
    nsim = 80,
    seed = 1000 + round(100 * weight_gravity)
  )
  predicted_cases <- predicted_infections

  list(
    prediction = predicted_cases,
    score = data.frame(
      candidate = sprintf("%s%% Facebook gravity", round(100 * weight_gravity)),
      weight_gravity = weight_gravity,
      log_RMSE = log_rmse(observed_cases, predicted_cases),
      total_predicted_cases = sum(predicted_cases)
    )
  )
}

weights <- c(0, 0.25, 0.5, 0.75, 1)
candidate_results <- lapply(weights, score_candidate)

scores <- do.call(rbind, lapply(candidate_results, `[[`, "score"))
scores <- scores[order(scores$log_RMSE), ]
scores$rank <- seq_len(nrow(scores))
rownames(scores) <- NULL
scores
```

The table ranks candidates for this simulated realization. The generating model
is known here, but stochastic variation means it need not always rank first.
Interpret the score together with the residual plots and consider sensitivity to
simulation variability before drawing conclusions.

## Inspect The Residual Pattern

A score gives the ranking, but the residual pattern shows what each model is
getting wrong. Here we compare the observed outbreak with the data-generating
model, the departure-diffusion-only alternative, and the best non-truth ensemble.

```{r}
prediction_by_weight <- setNames(
  lapply(candidate_results, `[[`, "prediction"),
  weights
)

best_ensemble_weight <- scores$weight_gravity[
  scores$weight_gravity > 0 & scores$weight_gravity < 1
][1]

comparison <- data.frame(
  region = names(observed_cases),
  observed = observed_cases,
  gravity = round(prediction_by_weight[["1"]]),
  departure_diffusion = round(prediction_by_weight[["0"]]),
  best_ensemble = round(prediction_by_weight[[as.character(best_ensemble_weight)]])
)

comparison <- comparison[order(comparison$observed, decreasing = TRUE), ]
comparison[1:12, ]
```

```{r, fig.width = 6.5, fig.height = 5.5}
prediction_plot_data <- rbind(
  data.frame(
    observed = as.numeric(observed_cases),
    predicted = as.numeric(prediction_by_weight[["1"]]),
    model = "Facebook gravity"
  ),
  data.frame(
    observed = as.numeric(observed_cases),
    predicted = as.numeric(prediction_by_weight[[as.character(best_ensemble_weight)]]),
    model = "Best intermediate ensemble"
  ),
  data.frame(
    observed = as.numeric(observed_cases),
    predicted = as.numeric(prediction_by_weight[["0"]]),
    model = "Facebook departure-diffusion"
  )
)

axis_labels <- function(x) format(round(x), big.mark = ",")

ggplot2::ggplot(
  prediction_plot_data,
  ggplot2::aes(observed, predicted, colour = model, shape = model)
) +
  ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey35") +
  ggplot2::geom_point(size = 2.4, alpha = 0.9) +
  ggplot2::scale_x_log10(labels = axis_labels) +
  ggplot2::scale_y_log10(labels = axis_labels) +
  ggplot2::scale_colour_manual(
    values = c(
      "Facebook gravity" = "black",
      "Best intermediate ensemble" = "#2b83ba",
      "Facebook departure-diffusion" = "#b2182b"
    )
  ) +
  ggplot2::labs(
    title = "Observed vs predicted outbreak size",
    subtitle = "Each point is a fictional region; dashed line shows perfect agreement",
    x = "Observed cumulative cases",
    y = "Predicted cumulative cases",
    colour = NULL,
    shape = NULL
  ) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    legend.position = "bottom",
    plot.title.position = "plot"
  )
```

This example demonstrates the 
workflow of exploring different candidate mobility models, fixing the transmission
model, scoring in some way (likely better than simple RMSE) against outbreak data, 
and using the residuals before deciding which mobility assumption is most 
useful for the epidemiological question.
