---
title: "Selecting a mobility model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Selecting a mobility model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "#>")
source(system.file("rmarkdown", "folding.R", package = "nomad"))
```

This walkthrough starts with a dataset, compares its fitted models, and predicts
movement for a common set of regions. You will finish with an explicit ensemble
and an interactive comparison. You need `nomad`; no private trip data are needed.
Long setup and plotting blocks can be expanded when you want to inspect the code.

```{r}
library(nomad)
```

## Find a relevant source

Start with the setting you want to study. Country alone is not enough: consider
how movement was measured, the administrative scale, and the population sampled.
Here we inspect the Zambia datasets. `has_fitted_models` distinguishes catalogue
entries from datasets you can actually use for prediction.

```{r}
mobility_table(country = "ZMB")[, c(
  "name", "aggregation", "has_fitted_models", "prediction_period_days", "distance_unit"
)]
```

The new Facebook models describe an average 21-day trip matrix. Collection dates
describe when observations were gathered, not the interval represented by each
prediction. The original Zambia models have unknown prediction periods, so we
will use the new dataset for this comparison.

## Compare fits to the same observations

A dataset can have several fitted model families. The following table shows their
errors and whether they support new locations. Lower RMSE means smaller typical
errors against that dataset's observed flows; it is not evidence that a model
will transfer successfully to another setting.

```{r}
candidates <- model_table(data_name = "zmb_fb_2025")
candidates[, c("name", "mobility_model", "RMSE", "DIC", "max_rhat", "supports_newdata")]
```

Inspect `max_rhat` alongside fit error. A high stored R-hat raises a convergence
concern; `rhat_above_1_1` is a review flag, not a claim that unflagged fits are
validated. Consult `summary(model)` for parameter-level diagnostics. Supplied
parameters are preserved, including flagged fits.

DIC was not calculated for these fits. Requesting DIC weights therefore gives an
error, including when only one selected model lacks DIC. We can instead request
RMSE weights because those statistics are available. Compare fit-derived weights
within one dataset; use deliberate manual weights for comparisons across sources.

```{r}
weights <- model_weights(data_name = "zmb_fb_2025", metric = "RMSE")
weights[, c("name", "RMSE", "weight")]
```

Weights are a starting point for sensitivity analysis. They do not account for
all differences between the source setting and your application.

## Inspect the source and diagnostics

We will compare exponential gravity with exponential departure-diffusion, both
fitted to the same Facebook dataset. The profile reports fitted distance units,
population scale and prediction period; the saved plot shows agreement with the
original observations even though private trip data are not distributed.

```{r}
gravity <- model_db$zmb_fb_2025_mod_grav_exp
departure <- model_db$zmb_fb_2025_mod_dd_exp
model_profile(gravity)
check(gravity, plots = FALSE)
```

```{r, fig.width=8, fig.height=4, fig.alt="Observed and fitted flows for Zambia exponential gravity"}
plot(gravity)
```

Look for systematic departures from agreement, not only the summary error.
A good source-data fit still needs the following checks before transfer.

## Prepare a common prediction setting

For a small example, take eight regions from the source covariates. Rows of `D`
identify origins, columns identify destinations, and `N` gives each region's
population. The example preserves those names so predictions can be aligned.
In a real application, replace these with your regions and population estimates.

```{r, fold_label="Show how to construct the common regions"}
source <- gravity$get_model()$data
regions <- rownames(source$D)[1:8]
newdata <- list(
  D = source$D[regions, regions],
  N = source$N_orig[regions]
)
```

These distances are in metres. Declaring `unit` makes conversion explicit; nomad
converts separately for each fitted model. The warnings about fewer regions or a
smaller population are useful here: this subset is smaller than the full source
setting. Transfer is supported, but its interpretation needs care.

```{r}
a <- predict(gravity, newdata = newdata, unit = "m", duration = 21)
b <- predict(departure, newdata = newdata, unit = "m", duration = 21)
round(a[1:4, 1:4])
```

Each entry is predicted trips from its row region to its column region over
21 days. It is not a probability or a count of unique travellers. A 42-day request
would double these values under a constant-rate assumption; it would not model
seasonality. Leave duration unset for an original Zambia model with an unknown
period: unscaled prediction warns, whereas duration conversion errors.

## Recognise unsupported requests

Basic and finite radiation models retain predictions for their exact original
regions and populations. They cannot transfer the stored matrix to the eight-region
subset above. Nomad rejects that request directly, including through an ensemble,
and explains how to choose a transferable model or recalculate radiation with
`mobility` when you have the necessary trip totals.

```{r}
radiation <- model_db$zmb_fb_2025_mod_rad_basic
original <- predict(radiation)
dim(original)
```

Changing the target country, spatial scale or population of a transferable model
produces advisory warnings, rather than this radiation capability error. Supply
`locations` as an `sf` object or longitude/latitude table to enable country checks.
No bundled model represents international travel. Check those warnings before
interpreting a cross-border prediction.

## Inspect uncertainty

A point prediction uses mean parameter estimates. With `nsim = 100`, a supported
model returns an array whose third dimension indexes simulated matrices. The
summary below describes variation in total and between-region movement; these
are parameter draws, not a guarantee of predictive calibration.

```{r}
draws <- predict(gravity, newdata = newdata, unit = "m", nsim = 100, seed = 42)
uncertainty_summary(draws, metrics = c("total", "between_region", "max"))
selected <- uncertainty_matrices(draws, metric = "between_region", probs = c(0.1, 0.5, 0.9))
attr(selected, "selected")
```

The selected matrices represent lower, middle and upper movement according to
that metric. They are complete draws, rather than a matrix made by taking an
independent quantile in each cell. Changing the metric may select different draws.

```{r, fig.width=7, fig.height=4, fig.alt="Lower, median and upper mobility matrices selected by between-region movement"}
plot_uncertainty(draws, metric = "between_region")
```

## Compare and combine predictions

Both matrices now describe the same named target regions and period. Their
difference is meaningful: positive entries below mean that gravity predicts more
trips than departure-diffusion. Nomad aligns reordered region names and rejects
matrices referring to different target regions.

```{r}
round(prediction_difference(a, b)[1:4, 1:4])
ensemble <- nomad_ensemble(list(gravity = gravity, departure = departure),
                           weights = c(gravity = 0.6, departure = 0.4))
combined <- predict(ensemble, newdata = newdata, unit = "m", duration = 21)
c(gravity = sum(a), departure = sum(b), ensemble = sum(combined))
```

These manual weights state our choice explicitly. Equal weights are also a valid
choice, but nomad will never substitute them when requested fit-based weights
cannot be calculated.

The slider below varies the weight of departure-diffusion (model B). The two
individual predictions stay fixed; the ensemble and its difference from gravity
update immediately. The same function works in an R notebook or saved HTML widget.

```{r}
compare_models(gravity, departure, newdata = newdata, unit = "m", duration = 21)
```

Carry several plausible mobility assumptions into your downstream analysis.
The outbreak vignette demonstrates how changing only mobility affects a simple
simulated epidemic, with the transmission assumptions held fixed.
